---
output: 
    rmarkdown::md_document:
        variant: gfm
---

# sitePath: an R package for detection of site fixation in molecular evolution

## Getting help

Post on Bioconductor [support site](https://support.bioconductor.org/) if having trouble using `sitePath`. Or open an [issue](https://github.com/wuaipinglab/sitePath/issues/new?assignees=&labels=&template=bug_report.md&title=) if a bug is found.

## Installation

[R programming language](https://cran.r-project.org/) >= 3.6.0 is required to use `sitePath`.

The stable release is available on [Bioconductor](https://bioconductor.org/packages/release/bioc/html/sitePath.html).
```r
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("sitePath")
```

The installation from [GitHub](https://github.com/wuaipinglab/sitePath/) is in experimental stage but gives the newest feature:
```r
BiocManager::install("wuaipinglab/sitePath")
```

## QuickStart

Both the phylogenetic tree and the matching multiple sequence alignment files are required.

```{r import_data, message=FALSE}
library(ape)
library(sitePath)

# The file names of your tree and multiple sequence alignment
tree_file <- system.file("extdata", "ZIKV.newick", package = "sitePath")
alignment_file <- system.file("extdata", "ZIKV.fasta", package = "sitePath")

tree <- read.tree(tree_file)
tree <- addMSA(tree, alignment_file, "fasta")
```

Use the `lineagePath` function to resolve major lineages (the choice of threshold really depends). Then use the `fixationSites` function to detect fixation sites.

```{r fixation_sites}
paths <- lineagePath(tree)

fixations <- fixationSites(paths)
print(fixations)
```
